---
title: "Practical1"
author: "Gia Bao hoang"
date: "2023-03-02"
output: html_document
---

# Set up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.width = 6, 
  fig.asp = 0.618, 
  out.width = "70%",
  fig.align = "center", 
  root.dir = "../"
)
pacman::p_load(tidyverse, gglm)
```

Load the data set
```{r}
data(Rubber, package = "MASS")
Rubber
```

# Plot

Plot hard against loss
```{r}
ggplot(Rubber,
       aes(hard, loss)
       ) + 
  geom_point()
```

Plot tens against loss
```{r}
ggplot(Rubber,
       aes(tens, loss)
       ) + 
  geom_point()
```

SIde by side comparision with trendline
```{r}
Rubber %>% 
  pivot_longer(-loss) %>%
  ggplot(aes(value, loss)) +
  geom_point() +
  facet_wrap(~name, scales = "free") + 
  geom_smooth(method = lm)

```

# Model

Fit the multiple linear regression
$$
  {E}(\texttt{loss}_i)=\beta_0+\beta_1\times\texttt{hard}_i+\beta_2\times\texttt{tens}_i
$$

```{r}
rubber_lm <- lm(loss ~ hard + tens, data = Rubber)
summary(rubber_lm)
```

Interpret the output:
- The least square estimate of $\beta_1$ is $\hat\beta_1=-6.5708$. The interpretation of this estimate is that an increase of 1 unit of hardness, with tensile strength kept constant, would reduce the rate of abrasion by 6.5708 grams per hour.
- The least square estimate of $\beta_2$ is $\hat\beta_2=-1.3743$. The interpretation of this estimate is that an increase of 1 unit of tensil strength, with hardness kept constant, would reduce the rate of abrasion by 1.3742 grams per hour

# Verify model

Verify the assumption of linear regression
```{r}
gglm(rubber_lm)
```
**Linearity**
 - Random scatter with a slight curvature near the end points -> satisfied
 
**Homocedasticity**
 - The spread about the zero line is roughly constant -> satisfied
 
**Normality**
 - There are departure from normality in the lower tail. However, the rest of 
 the data is closed to normally distributed -> satisfied

# Prediction
 
Predict loss for the following points 
```{r}
n_pts <- tibble(
  hard = c(50, 65),
  tens = c(200, 190)
)
n_pts
```
 
```{r}
predict(rubber_lm)
predict(rubber_lm, newdata = n_pts)
```
 Calculate 95% CI and PI for the new points
 
```{r}
predict(rubber_lm, newdata = n_pts, interval = "confidence")
predict(rubber_lm, newdata = n_pts, interval = "prediction")
```
# Matrix

The design matrix of the model
```{r}
x <- model.matrix(rubber_lm)
x
```

Assign the response variable `loss` to a variable Y
```{r}
Y <- Rubber$loss
Y
```

Calculate $\hat\beta=(X^TX)^{-1}X^TY$
```{r}
beta_hat <- solve(t(x) %*% x) %*% t(x) %*% Y
beta_hat

```

Calculate $\hat{\boldsymbol{\eta}}=X\hat{\boldsymbol{\beta}}$

```{r}
eta <- x %*% beta_hat
eta
```

Calculate the residual variance, $s^2_e$ directly from the observed and fitted values
 
```{r}
n <- nrow(Rubber)
se <- sqrt(sum((Y - eta)^2) / (n - 3))
se
```

Calculate the estimated variance matrix for $\hat{\boldsymbol{\beta}}$ using
$$
  (X^TX)^{-1} \times s^2_e
$$
Compare this to the result of the built-in calculation vcov()
```{r}
vcov(rubber_lm)
res <- solve(t(x) %*% x) * se^2
res
```
 
 
 
 
 
 